Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-10-03 article/ms.tex
reorder benchmark method

We introduce and evaluate two search strategies that make use of the objective function in Equation \ref{eq:objective_function}. These search strategies are compared against an existing state-of-the-art method, which we describe in Section \ref{sec:experimental-design}.

\subsection{The Message-Breaking Method} \label{sec:mbm}
The first strategy, \emph{The Message-Breaking Method}, does not make any predictions on
the message length of future mixtures. Instead it starts from 
a $K=1$ mixture and iteratively splits the component with the largest message length until 
the total message length no longer improves (Algorithm~\ref{alg:MBM}). We evaluate the message length of each component
by re-ordering Equation \ref{eq:objective_function} and weighting the negative log-likelihood
$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ with entries
\begin{eqnarray}
	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
\end{eqnarray}
\noindent{}such that the message length of the $k$th component is
\begin{eqnarray}
	I_k &=& -\sum_{n=1}^{N}\mathcal{R}_{nk}\log{\weight_{k}f(\data_{n}|\vecmean_k,\veccov_k)} -\frac{(D + 2)}{2}\log{|\cov_k|} \nonumber \\
		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right]
\end{eqnarray}
\noindent{}and the total message length remains the sum of the message lengths of the components
\begin{eqnarray}
	I(\vec\data|\vec\psi, N, D) = \sum_{k=1}^{K}I_{k} \quad .
\end{eqnarray}

The component-wise calculation of the message length is performed as part of the 
expectation step of the expectation-maximisation algorithm, adding negligible computational overhead.

@@

One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. However, as we show in Section \ref{sec:results}, this hypothetical situation was not observed in any of the $\todo{5000}$ generated datasets.

@@

\subsection{Benchmark Method}
We use the search algorithm proposed \citep{kasarapu2015minimum} as a benchmark for comparison, which has been shown to be the state-of-the-art.
% in what sense has this been shown?
~\cite{kasarapu2015minimum}. 
\begin{algorithm}[H]
\caption{The benchmark method~\cite{kasarapu2015minimum}.}
\label{alg:kasarapu}
\begin{algorithmic}[1]
\State {\bf \sc input:} $K$ \Comment{Number of components.}
\State current $\leftarrow$ one component mixture
\While{TRUE}
\State components $\leftarrow$ current
  \For{$i\in K$}\label{line:splitBegin}\Comment{Exhaustively split all components}
    \State splits[i]$\leftarrow$ \Call{Split}{current,components[i]}
  \EndFor
\State  bestSplit$\leftarrow$\Call{Best}{splits}\label{line:splitEnd} \Comment{Record the best split.}
\If{K>1}
  \For{$i\in K$}\label{line:deleteBegin}\Comment{Exhaustively delete all components}
    \State deletes[i]$\leftarrow$ \Call{Delete}{current,components[i]}
  \EndFor
\State  bestDelete$\leftarrow$\Call{Best}{deletes} \label{line:deleteEnd}\Comment{Record the best delete.}
\EndIf
  \For{$i\in K$}\label{line:MergeBegin}\Comment{Exhaustively merge all components}
    \State j$\leftarrow$ \Call{closestComponent}{i}
    \State merges[i]$\leftarrow$ \Call{Merge}{merges}
  \EndFor
\State  bestMerge$\leftarrow$\Call{Best}{merges} \label{line:MergeEnd}\Comment{Record the best merge.}
\State bestPerturbation $\leftarrow$ \Call{Best}{bestSplit, bestDelete,bestMerge} \Comment{Select the best perturbation.}
\State $\Delta I \leftarrow$ \Call{MessageLength}{bestPerturbation}-\Call{MessageLength}{current} \Comment{Check for improvement.}
\If{$\Delta I < 0$}
\State current $\leftarrow$ bestPerturbation
\State {\bf \sc Continue}
\EndIf
\State {\bf \sc Break} \label{line:end}
\EndWhile
\State {\bf \sc output:} current
\end{algorithmic}
\end{algorithm}

The method is presented in Algorithm~\ref{alg:kasarapu}. It starts with one component and iteratively splits (lines \ref{line:splitBegin}-\ref{line:splitEnd}), deletes (lines \ref{line:deleteBegin}-\ref{line:deleteEnd}), and merges (lines \ref{line:MergeBegin}-\ref{line:MergeEnd}) components until no improvements in the message length is observed (line~\ref{line:end}).

\subsection{Dataset}

\alda{Andy, can you please include the details?}
\paragraph{Simulations}

\paragraph{GA data}

\subsection{Criteria for Comparison}

\section{Results}
\section{References}
\bibliographystyle{elsarticle-num}
\bibliography{mmlbibliography}

@@

>>>>

We introduce and evaluate two search strategies that make use of the objective function in Equation \ref{eq:objective_function}. These strategies are benchmarked against an existing state-of-the-art search method, which we have adapted to use the same objective function in Equation~\ref{eq:objective_function} in order to facilitate a fair comparison.

\subsection{Benchmark Method}

We use the search algorithm proposed \citep{kasarapu2015minimum} as a benchmark for comparison. The method is presented in Algorithm \ref{alg:kasarapu}. It starts with a single component mixture and iteratively splits (lines \ref{line:splitBegin}-\ref{line:splitEnd}), deletes (lines \ref{line:deleteBegin}-\ref{line:deleteEnd}), and merges (lines \ref{line:MergeBegin}-\ref{line:MergeEnd}) components until no improvements in the message length is observed (line~\ref{line:end}). This is a greedy approach, which is advantageous in that it is guaranteed to only accept perturbations that minimise the message length, but that guarantee comes with a computational cost. 
% in what sense has this been shown?
\begin{algorithm}[H]
\caption{The benchmark method~\cite{kasarapu2015minimum}.}
\label{alg:kasarapu}
{\bf Input:} $\vec\data$ \\
{\bf Output:} $\vec\psi$
\begin{algorithmic}[1]
\State current $\leftarrow$ one component mixture
\While{TRUE}
\State components $\leftarrow$ current
  \For{$i\in K$}\label{line:splitBegin}\Comment{Exhaustively split all components}
    \State splits[i]$\leftarrow$ \Call{Split}{current,components[i]}
  \EndFor
\State  bestSplit$\leftarrow$\Call{Best}{splits}\label{line:splitEnd} \Comment{Record the best split.}
\If{K>1}
  \For{$i\in K$}\label{line:deleteBegin}\Comment{Exhaustively delete all components}
    \State deletes[i]$\leftarrow$ \Call{Delete}{current,components[i]}
  \EndFor
\State  bestDelete$\leftarrow$\Call{Best}{deletes} \label{line:deleteEnd}\Comment{Record the best delete.}
\EndIf
  \For{$i\in K$}\label{line:MergeBegin}\Comment{Exhaustively merge all components}
    \State j$\leftarrow$ \Call{closestComponent}{i}
    \State merges[i]$\leftarrow$ \Call{Merge}{merges}
  \EndFor
\State  bestMerge$\leftarrow$\Call{Best}{merges} \label{line:MergeEnd}\Comment{Record the best merge.}
\State bestPerturbation $\leftarrow$ \Call{Best}{bestSplit, bestDelete,bestMerge} \Comment{Select the best perturbation.}
\State $\Delta I \leftarrow$ \Call{MessageLength}{bestPerturbation}-\Call{MessageLength}{current} \Comment{Check for improvement.}
\If{$\Delta I < 0$}
\State current $\leftarrow$ bestPerturbation
\State {\bf \sc Continue}
\EndIf
\State {\bf \sc Break} \label{line:end}
\EndWhile
\end{algorithmic}
\end{algorithm}

\subsection{The Message-Breaking Method} \label{sec:mbm}
The simplest strategy we introduce here, \emph{The Message-Breaking Method}, does not use any predictions of the message length of future mixtures. Instead it starts from 
a $K=1$ mixture and iteratively splits the component with the largest message length until 
the total message length no longer improves (Algorithm~\ref{alg:MBM}). This approach is related to the benchmark method in that we iteratively move from a $K=1$ mixture, but we only split components (instead of splitting, merging, and deleting all components). Similarly, in \emph{The Message-Breaking Method} we only split the component with the largest message length instead of exhaustively trying to split all components at every mixture.

We evaluate the message length of each component by rearranging Equation \ref{eq:objective_function} and multiplying the negative log-likelihood
$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ that has entries
\begin{eqnarray}
	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
\end{eqnarray}
\noindent{}such that the message length of the $k$th component is
\begin{eqnarray}
	I_k &=& -\sum_{n=1}^{N}\mathcal{R}_{nk}\log{\weight_{k}f(\data_{n}|\vecmean_k,\veccov_k)} -\frac{(D + 2)}{2}\log{|\cov_k|} \nonumber \\
		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right]
\end{eqnarray}
\noindent{}and the total message length remains the sum of the message lengths of the components
\begin{eqnarray}
	I(\vec\data|\vec\psi, N, D) = \sum_{k=1}^{K}I_{k} \quad .
\end{eqnarray}

The component-wise calculation of the message length is performed as part of the 
expectation step of the expectation-maximisation algorithm, adding negligible computational overhead to the algorithm.

@@

\andy{Should we move/remove the following paragraph?}
One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. However, as we show in Section \ref{sec:results}, this hypothetical situation was not observed in any of the $\todo{5000}$ generated datasets.

@@

\subsection{Generated Datasets}

\todo{How we generated data sets}

\todo{Range of generated data sets}

\todo{Figure: predictions for future mixtures}

\todo{Figure: computational cost of various methods}

\todo{Figure: corner plot showing chemical abundance data}

\todo{Figure: corner plot showing chemical abundance data, coloured by their responsibility}

\todo{Figures showing the concentration of chemically-identified data}

\paragraph{Simulations}

\paragraph{Criteria for Comparison}

\subsection{Galactic Archaeology data}

\section{Results}

\todo{Convergence criteria: how often do we reach the `true' solution?}

\todo{What is the variance in the results of our method?}

\todo{Speed comparisons}

\todo{What do we find for the chemical abundance data}

\todo{Recovering known star clusters that are still gravitationally bound?}

\todo{Scientific insight from identifying groups in these data?}

\todo{Estimate of the number of star-forming sites in the galaxy?}

\section{References}
\bibliographystyle{elsarticle-num}
\bibliography{mmlbibliography}

@@

\subsubsection{Search Strategy Implementation}

The astute reader will recognise that each search `strategy' we have described
is implemented in the \texttt{gmmmml} codebase as a set of `policies' that govern
different behaviour of the search:
\begin{itemize}
\item \emph{Initialisation}: how many mixtures should be initialised, and how?
\item \emph{Prediction}: when should we make predictions about the message length of future mixtures, and how far ahead should we make predictions?
\item \emph{Movement}: \emph{what} should the next $K$ be?
\item \emph{Repartition}: \emph{how} should we move to the next $K$? Should we repartition an existing mixture, initialise a new mixture, or should we make a probabilistic decision?
\item \emph{Convergence}: when should we stop searching?
\end{itemize}

As an example, the \emph{Message-Breaking Method} can equivalently be described
by a set of policies that: 
	\emph{initialise} a single mixture $K = 1$; 
	makes no \emph{predictions};
	\emph{moves} a single step towards the minimum message length;
	\emph{repartitions} the closest existing mixture by splitting the component with the maximum message length; and 
	describes \emph{convergence} at the point when the message length of the current mixture is worse than the previous mixture. 
This policy-strategy factorisation allows for the trivial implementation of flexible search strategies with complex behaviour. We encourage the development of new search strategies.
